---
title: "McElreath _ Ch 15b"
output: html_notebook
---

# Missing data  
(section 15.2)  
Corresponds to Lecture 20, starting 32.21.

Common approaches:  
Drop missing data, only use complete cases - lose a lot of data and can create confounds  

Imputation
  substituting mean of column is very bad idea - model treats as if model is known, when in fact there is error  
  Multiple imputation - frequentist approach that works v well
  Bayesian imputation  
  
Why impute? Missingness can be a confound.  

## Types of missingness  

### Missing completely at random  
No confound. Don't need to impute, though it will improve precision.   
Missingness not affected by anything else we know.  

### Missing at random  
E.g. with body mass and primates: body mass, brain size, and milk energy.  This would be the case if body mass affected whether brain size measured.  
Need to close the back door; can condition on M and then impute.  

So this means that there is conditional independence of the missingness mechanism. 
If we delete incomplete cases, we obscure causation.  

## Missing not at random  
Missing more likely for specific values of dependent variable.  
Creates backdoor that can't be closed.  
Can also arise through unobserved variables. 
Only hope is to model missingness mechanism.  

Draw a DAG. I found syntax from earlier version of book to make a v fancy DAG in ggplot, so will try that out. Here attempting to make fig 15.4 with 4 panels.   
see https://bookdown.org/content/4857/the-haunted-dag-the-causal-terror.html

```{r missingnessDAG}
require(rethinking)
library(ggdag)
library(tidyverse)

dag_coords <-
  tibble(name = c("S", "D", "H", "I"),
         x    = c(1, 1, 2, 2),
         y    = c(2, 1, 2, 1))
ggtit <- 'A: completely at random'
g<-dagify(I ~ H,
       I ~ D,
       H ~ S,
       coords = dag_coords) %>%
  ggplot(aes(x = x, y = y, xend = xend, yend = yend))

formatdag <- function(g,ggtit){ #make into function so can reuse for other plots
g+  geom_dag_point(aes(color = name == "H"),
                 alpha = 1/2, size = 6.5, show.legend = F) +
  geom_point(x = 2, y = 1, 
             size = 6.5, shape = 1, stroke = 1, color = "orange") +
  ggtitle(ggtit) +
  geom_dag_text(color = "black") +
  geom_dag_edges() +
  scale_color_manual(values = c("steelblue", "orange")) +
  scale_x_continuous(NULL, breaks = NULL, expand = c(.1, .1)) +
  scale_y_continuous(NULL, breaks = NULL, expand = c(.1, .1))}

formatdag(g,ggtit)

#2nd plot
ggtit <- 'B. Missing at random'
g<-dagify(I ~ H,
       I ~ D,
       H ~ S,
       D ~ S,
       coords = dag_coords) %>%
  ggplot(aes(x = x, y = y, xend = xend, yend = yend))
formatdag(g,ggtit)

#4th plot
ggtit <- 'D. Missing not at random'
g<-dagify(I ~ H,
       I ~ D,
       H ~ S,
       D ~ H,
       coords = dag_coords) %>%
  ggplot(aes(x = x, y = y, xend = xend, yend = yend))
formatdag(g,ggtit)

#3rd plot





gg_fancy_dag <- function(d, x = 1, y = 1, circle = "X") {
  
  d %>% 
    ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
    geom_dag_point(aes(color = name == circle),
                   alpha = 1/2, size = 6.5, show.legend = F) +
    ggtitle(ggtit)+
    geom_point(x = x, y = y, 
               size = 6.5, shape = 1, stroke = 1, color = "orange") +
    geom_dag_text(color = "black") +
    geom_dag_edges() + 
    scale_color_manual(values = c("steelblue", "orange")) +
    theme_dag()
  
}

ggtit <- 'C. Alternative version of Missing at random'
dag_coords2 <-
  tibble(name = c("S", "D", "H", "I","X"),
         x    = c(1, 1, 3, 3,2),
         y    = c(3, 1, 3, 1,2))
g<-dagify(I ~ H,
       I ~ D,
       H ~ S,
       D ~ X,
       H ~ X,
       coords = dag_coords2)
gg_fancy_dag(g, x = 1, y = 1, circle = "X")
```
D is dog  
H is homework  
I is observed homework (ie with missing data)  
S is student who studies  
X is noisy home (unobserved)  

I don't this example works terribly well....  Not least because dog eating homework is not same as missing data?

But I am very interested in making analogy from this to ascertainment bias. That's where data is missing in a biased fashion. E.g. you only recruit children to a study if they have problems.  

Note re simulations: even when we cannot completely eliminate impact of missing data w might be able to show through simulation that the impact is small.

Here are some simulations.  First missing completely at random.

```{r code15.8}
N <- 100
S <- rnorm(N) #amount of studying
H <- rbinom(N, size=10, inv_logit(S)) #grading of homework on 10 pt scale

#reminder of inv logit function
plot(S,inv_logit(S))

plot(S,H,xlab='studying',ylab='Homework score') #more studying, higher score

#code15.9
#Simulate random dog eating
 D <- rbern(N)
 head(D) #just zeros and ones for each observation
 Hm <- H
 Hm[D==1]<-NA #turn ones into NA
 head(Hm)

table(D,H) #D and H are unrelated: D is random
```

From DAG can see that H is independent of D because I is a collider.  

Missingness doesn't change overall distribution of H scores. Removes data so estimation is less efficient. 

Now model B.  Studying influences whether dog eats homework.  

```{r code15.10}
D <- ifelse(S>0,1,0)
Hm <- H
Hm[D==1] <- NA
plot(S,H,xlab='studying',ylab='Homework score') #more studying, higher score
plot(S,Hm,xlab='studying',ylab='Homework score, exclude missing') #more studying, higher score

```

So this is a fairly extreme simulation where dog eats homework of all those who study above average.  

Creates noncausal backdoor path   H* <- D <- S.
Can close by conditioning on S.

(? how well does that work if S range completely truncated?)  

Figure 15.4 c shows a noncausal variable that affects H and D, but it is an unobserved variable, X - e.g. noise in the home.
If we regress Hm on S we have Hm <- D <- X -> H  
Code to simulate this:

```{r code15.11}
set.seed(501)
N <- 1000
X <- rnorm(N)
S <- rnorm(N)
H <- rbinom(H, size=10,inv_logit(2+S-2*X))
D <- ifelse(X>1,1,0)
Hm <- H
Hm[D==1]<- NA

```


now run model
```{r code15.12}



```

  










